Light inextensible strings under tension in the Schwarzschild geometry

### 2020-04-10 string_physics_teacher_revision.tex
rephrase for TPT, issue #20

Of course, no real string is perfectly weightless or inextensible.
With tension~$T$, gravitational acceleration $g$ and (proper) mass per
(proper) unit length~$\mu$, then a string will behave as though light
on lengthscales $L\lesssim T/\left(\mu g\right)$.  Using
$c^4/\left(4GM\right)$ for the strength of gravity at the event
horizon of a black hole with mass $M$ gives $L\lesssim TM/\left(\mu
c^4\right)$.  For the string to behave as though light on lengthscales
comparable with the Schwarzschild radius would require $T\gtrsim\mu
c^2$.  Exceeding this criterion would violate the dominant energy
condition (and in this context we note that transverse waves would
propagate at speed~$\sqrt{T/\mu}\sim c$).  Nevertheless, the
idealization of a string as light and inextensible is an interesting
and informative limit: Rindler~\cite{rindler}, for example, considers
lowering a heavy object to the event horizon on the end of such a
string to illustrate energetic arguments in GR.

@@

\noindent Here I consider the spatial component~$ds^2=
dr^2/\left(1-2M/r\right) + r^2d\phi^2$, confined to equatorial
points~($\theta=\pi/2$); we understand~$r\geqslant 2M$ throughout.  A
circular string at constant~$r$ will, trivially, have proper
length~$2\pi r$.

Considering first the radial component we may calculate the
proper length $L\left(r_1,r_2\right)$ of a string stretching
from~$r=r_1\geqslant 2M$ to~$r_2\geqslant r_1$ at constant $\theta,\phi$:

% \begin{equation}\label{radial_string_length_properunits}
%   \int_{r=r_1}^{r_2}\frac{dr}{\sqrt{1-1/r}}=
%   \left.
%   \sqrt{r(r-1)} +\frac{1}{2}\log\left(
%   \frac{\sqrt{r}+\sqrt{r-1}}{\sqrt{r}-\sqrt{r-1}}\right)
%   \right|_{r_1}^{r_2}
%   \end{equation}

@@

%  \begin{equation}\label{ell_properunits}
%    \ell=\int_{r=1}^{r=1+w}ds
%    =2\sqrt{w} + \frac{1}{3}w^{3/2} + O(w^{5/2})
%  \end{equation}

@@

% \begin{equation}\label{asymptotic_ell_properunits}
%   \ell = w +  \frac{1}{2}\log w + \frac{1}{2} + \log(2)  +  \frac{1}{8}w^{-1} + O(w^{-2}),
% \end{equation}
% 
\begin{equation}\label{asymptotic_ell}
  \ell = 2M\left(w +  \frac{1}{2}\log w + \frac{1}{2} + \log(2)  +  \frac{1}{8}w^{-1} + O(w^{-2})\right),
\end{equation}

@@

@@

% \begin{equation}
% %  \int_{p_1}^{p_2}\sqrt{\frac{\left(r'\right)^2}{1-2M/r} + r^2}d\phi=
%   \int_{p_1}^{p_2}\left(\frac{\left(r'\right)^2}{1-1/r} + r^2\right)^\frac{1}{2}d\phi=
%   \int_{p_1}^{p_2}F\left(r,r'\right)d\phi
% \end{equation}

@@

\noindent Solutions to equation~\ref{rdashdash} correspond to a taut
string confined to Flamm's paraboloid~\cite{flamm1916}; see
Figure~\ref{flamm}.  It is interesting to compare
equation~\ref{rdashdash} with the corresponding equation for null
geodesics, which is usually given~\cite{wald} in terms of $u=2M/r$ as
$u''=3u^2/2-u$, but is equivalent
to~$r''\left(\phi\right)=r-3M+2\left(r'\right)^2/r$.  This shows that
light, being sensitive to the time component of the metric, behaves
differently from string, which is not (at this point it might be worth
``reminding'' students that a straight line in flat space,
$r=r_0\sec\left(\phi-\phi_0\right)$, obeys
$r''=r+2\left(r'\right)^2/r$).

There does not appear to be a simple analytical solution to
equation~\ref{rdashdash}; but it is a straightforward second order
ODE, amenable to numerical solution and no harder than the equivalent
for null geodesics, a standard homework problem.  Here we consider
only strings exterior to the event horizon, $r>2M$, which have a
well-defined point of closest approach to the singularity.  Because
the system is invariant under~$\phi\longrightarrow-\phi$, the string
must be symmetrical about this point.  If $\phi'$ is small, then a
string follows $r''=r-2M$, while tangential null geodesics follow
$r''=r-3M$.  Thus a taut string would {\em look} curved to a local
observer, having a (visual) radius of curvature of
$r_0^2/M\sqrt{1-2M/r_0}$ (the extra term comes from the fact that
appearance is driven by local $ds$, not Schwarzschild $r$).  At the
photon sphere $r_0=3M$, such a string would appear to have a radius
of~$\sqrt{27}M$.  For a solar mass black hole this would be about
$15\unit{km}$, a readily understandable result.

\section{Numerical results}

Here I solve equation~\ref{rdashdash} numerically, using the R
programming language~\cite{rcore2019} and the \verb+deSolve+ package
for ordinary differential equations~\cite{soetart2010}.  First we
consider strings for which $r>2M$ everywhere (that is, completely
exterior strings).
Figure~\ref{closest_approach_non_self_intersecting} shows a sequence
of non-self-intersecting numerical solutions to
equation~\ref{rdashdash}, rotated so that the tangential point of
closest approach occurs at~$\phi=0$.
Figure~\ref{closest_approach_self_intersecting} shows the
corresponding diagram for self-interesecting strings
(Figure~\ref{light_closest_approach} shows equivalent null geodesics
for comparison).  A bisection technique reveals that if the closest
approach is greater than a critical value of about~$\simeq 1.0761\cdot
2M$, the string does not self-intersect; if the closest approach is
less than this value, then the string crosses itself at~$\phi=\pi/2$.

@@

The concept of ``straight line'' is problematic in general relativity.
Although photon paths provide a definition of straightness, their
world lines are sensitive to the time component of the metric.  Light
taut strings respond only to the spatial components of the metric, and
their configuration is arguably a more intuitive and helpful way to
think about straightness.  Using numerical methods to compute string
configuration is suitable for an undergraduate homework problem.

@@

\begin{figure}[p] % created by "closest_approach2.R"
\centering
\includegraphics[width=\linewidth]{closest_approach_selfintersecting.pdf}
\caption{Light inextensible strings under tension close to a black
  hole.  Here self-intersecting strings are shown for $\phi\geqslant
  0$.  Closest approach to the event horizon occurs at $\phi=0$ (3
  o'clock) and $r/2M\in\left(1,1.0761\right)$; strings are symmetrical
  about $\phi=0$ but for clarity only $\phi\geqslant 0$ is shown}
\label{closest_approach_self_intersecting}
\end{figure}

>>>>

Of course, no real string is perfectly weightless or inextensible.
Students could use dimensional analysis to assess the accuracy of such
approximations.  With tension~$T$, gravitational acceleration $g$ and
(proper) mass per (proper) unit length~$\mu$, then a string will
behave as though light on lengthscales $L\lesssim T/\left(\mu
g\right)$.  Using $c^4/\left(4GM\right)$ for the strength of gravity
at the event horizon of a black hole with mass $M$ gives $L\lesssim
TM/\left(\mu c^4\right)$.  For the string to behave as though light on
lengthscales comparable with the Schwarzschild radius would require
$T\gtrsim\mu c^2$.  Exceeding this criterion would violate the
dominant energy condition (and in this context we might remind
students that transverse waves would propagate at
speed~$\sqrt{T/\mu}\sim c$).  Nevertheless, the idealization of a
string as light and inextensible is an interesting and instructive
limit: Rindler~\cite{rindler}, for example, considers lowering a heavy
object to the event horizon on the end of such a string to illustrate
energetic arguments in GR.

@@

Many undergraduate courses will present the Schwarzschild metric {\em
  ex nihilo}, in which case its pedagogical value would be to
illustrate non-Newtonian features of GR such as time dilation.  The
most striking of these are temporal: suppose pulses of light are
emitted from a distant source [such as a pulsar] with interval $t_P$;
the coefficient of $dt^2$ implies that a student, stationary at
distance $r>2M$ from the singularity will perceive these pulses to
have an interval of~$t_P/\sqrt{1-2M/r} < t_P$ [it might be worth
  emphasising that coordinate ``$t$'' coincides with Lorentzian time
  for an observer at infinity].  Further, if our student emits pulses
at (local) interval $t_L$, then the distant observer would see these
pulses with an interval of $t_L\sqrt{1-2M/r}$, and this approaches
infinity as $r\longrightarrow 2M$ and the distant observer sees the
student ``freezing''.  Nevertheless, the free-fall time from the Event
Horizon to the singularity is finite.

We also see a failure of classical geometry: the coordinate $r$ has
the defining feature that the circumference of a circle of radius $r$
has circumference $2\pi r$; but radial distances are not Euclidean.
To investigate this phenomenon, we consider the spatial
component~$ds^2= dr^2/\left(1-2M/r\right) + r^2d\phi^2$, confined to
equatorial points~($\theta=\pi/2$); we understand~$r\geqslant 2M$
throughout.

Students might need to be reminded of the concepts of proper time and
proper length: these are quantities measured {\em locally}.  In
particular, we note that proper length of an object is the length as
measured by a local observer at rest with respect to the object.  The
proper length of a string is the length measured by an inchworm as it
crawls along the string.  A circular string at constant~$r$ will,
trivially, have proper length~$2\pi r$.  Considering first the radial
component we may calculate the proper length $L\left(r_1,r_2\right)$
of a string stretching from~$r=r_1\geqslant 2M$ to~$r_2\geqslant r_1$
at constant $\theta,\phi$:

@@

@@

\begin{equation}\label{asymptotic_ell}
  \ell = 2M\left(w +  \frac{1}{2}\log w + \frac{1}{2} + \log(2)  +  \frac{1}{8}w^{-1} + O(w^{-2})\right),
\end{equation}

@@

\section{Three types of straight line}

There are at least three types of ``straight line'' to consider when
studying general relativity.  In the context of
equation~\ref{schwarzschild} we have:

\begin{enumerate}
\item null geodesics: paths for which $ds^2=0$ at all points.  Light
  travels on null geodesics and is arguably the ``straightest'' of
  straight lines.
\item timelike geodesics: paths for which $ds^2>0$ everywhere, and for
  which $\int\sqrt{ds^2}$ is a minimum.  These are the paths that
  objects in freefall follow.
\item taut string: paths for which $dt=0$, thereby considering the
  configuration of a string.  If the proper length
  $\int\sqrt{dr^2+r^2d\theta^2}$ is a minimum, the string is taut.
\end{enumerate}

The path of a taut string seems to be a natural and intuitive
interpretation of straight line, with immediate everyday
interpretation, yet has been neglected in the pedagogical literature.
Nontrivial taut strings require analysis of nonradial paths, which I
consider below.

@@

@@

\noindent Solutions to equation~\ref{rdashdash} correspond to a taut
string confined to Flamm's paraboloid~\cite{flamm1916}; see
Figure~\ref{flamm}.  At this point it might be worth ``reminding''
students that a straight line in flat space,
$r=r_0\sec\left(\phi-\phi_0\right)$, obeys
$r''=r+2\left(r'\right)^2/r$, and that equation~\ref{rdashdash}
reduces to this in the limit $r/2M\longrightarrow\infty$.

It is interesting to compare equation~\ref{rdashdash} with the
corresponding equation for null geodesics, which is usually
given~\cite{wald} in terms of $u=2M/r$ as $u''=3u^2/2-u$, but is
equivalent to~$r''\left(\phi\right)=r-3M+2\left(r'\right)^2/r$.  This
shows that light, being sensitive to the time component of the metric,
behaves differently from string, which is not.

Here we consider only strings exterior to the event horizon, $r>2M$,
which have a well-defined point of closest approach to the
singularity.  Because the system is invariant
under~$\phi\longrightarrow-\phi$, the string must be symmetrical about
this point.  If $\phi'$ is small, then a string follows $r''=r-2M$,
while tangential null geodesics follow $r''=r-3M$.  Thus a taut string
would {\em look} curved to a local observer, having a (visual) radius
of curvature of $r_0^2/M\sqrt{1-2M/r_0}$ (the extra term comes from
the fact that appearance is driven by local $ds$, not Schwarzschild
$r$).  At the photon sphere $r_0=3M$, such a string would appear to
have a radius of~$\sqrt{27}M$.  For a solar mass black hole this would
be about $15.4\unit{km}$, a readily understandable result.  Of course,
real cotton thread would break here, but the deviation of tangential
light from a taut string's path would be about $0.13\unit{mm}$,
certainly detectable by the naked eye.
\section{Numerical results}

There does not appear to be a simple analytical solution to
equation~\ref{rdashdash} and numerical methods are needed.  In an
educational context, we might use Euler's method to illustrate the
concepts and find $\left(r,r'\right)$---that is, a point in a
two-dimensional phase space---as a function of independent variable
$\phi$.  We specify initial conditions $(r_0,r'_0)$ at $\phi_0$ and
repeatedly increment $\phi$ by a small amount $\delta\phi$:

\begin{eqnarray}
\phi&\longrightarrow&\phi+\delta\phi\nonumber\\
r   &\longrightarrow&r+r'\delta\phi\nonumber\\
r'  &\longrightarrow&r' + r''\delta\phi\qquad\mbox{where $r''$ is from equation~\ref{rdashdash}.}
\end{eqnarray}

Of course, Euler's method is numerically unstable and inefficient;
here use the R programming language~\cite{rcore2019} and the
\verb+deSolve+ package for ordinary differential
equations~\cite{soetart2010} using the standard \verb+lsoda+ solver.
The equations typically imply
$\displaystyle\lim_{\phi\longrightarrow\pm\phi_C}=\infty$ for some
critical value $\phi_C$ of $\phi$ [observe $\phi_C=\pi/2$ for flat
  space] and this makes numerical analysis challenging.

First we consider strings for which $r>2M$ everywhere (that is,
completely exterior strings).
Figure~\ref{closest_approach_non_self_intersecting} shows a sequence
of non-self-intersecting numerical solutions to
equation~\ref{rdashdash}, rotated so that the tangential point of
closest approach occurs at~$\phi=0$.

Further analysis reveals something unexpected: for some values of
initial conditions, the string self-intersects even though it is under
tension everywhere and cannot support any nonzero bending moment.
Figure~\ref{closest_approach_self_intersecting} shows some
self-interesecting strings (Figure~\ref{light_closest_approach} shows
equivalent null geodesics for comparison).  A bisection technique
reveals that if the closest approach is greater than a critical value
of about~$\simeq 1.0761\cdot 2M$, the string does not self-intersect;
if the closest approach is less than this value, then the string
crosses itself at~$\phi=\pi/2$.

@@

The concept of ``straight line'' is problematic in general relativity.
Although photon paths provide a definition of straightness, their
world lines are sensitive to the time component of the metric.  Light
taut strings respond only to the spatial components of the metric, and
their configuration is arguably a more intuitive and helpful way to
think about straightness.  Using numerical methods to compute string
configuration is suitable for an undergraduate homework problem; the
difficult part is specifying sensible initial conditions.

The code producing the figures is available at github,
\url{http://github.com/RobinHankin/string}.

@@

\begin{figure}[p] % created by "closest_approach3.R"
\centering
\includegraphics[width=\linewidth]{closest_approach_selfintersecting_onegreen.pdf}

\caption{Light inextensible strings under tension close to a black
  hole.  Here, self-intersecting strings are shown for $\phi\geqslant
  0$ for clarity, but a single string is shown in green for
  $-3.69\leqslant\phi\leqslant 3.69$, making the self-intersection
  evident (all strings in the diagram, whether blue or green,
  self-intersect at $\phi=\pm\pi$, even if the point of intersection
  is far to the right).  Closest approach to the event horizon occurs
  at $\phi=0$ (3 o'clock) and $r/2M\in\left(1,1.0761\right)$; all
  strings are symmetrical about $\phi=0$}
\label{closest_approach_self_intersecting}
\end{figure}
